Spectroscopic confirmation of very-wide stellar binaries and large-separation comoving pairs from \textsl{Gaia DR1}

### 2017-10-02 paper/comoving-followup.tex
für
placeholder section on mixture model

We thank James Davenport (WWU) for releasing his open source spectroscopic
reduction code, \package{PyDIS}, which served as inspiration for the pipeline
used in this work.
This work is based on observations obtained at the MDM Observatory, operated by
Dartmouth College, Columbia University, Ohio State University, Ohio University,
and the University of Michigan.
This work has made use of data from the European Space Agency (ESA)
mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by
the {\it Gaia} Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding
for the DPAC has been provided by national institutions, in particular
the institutions participating in the {\it Gaia} Multilateral Agreement.
This research has made use of the SIMBAD database, operated at CDS, Strasbourg,
France.
This project was developed in part at the 2017 Heidelberg Gaia Sprint, hosted by
the Max-Planck-Institut für Astronomie, Heidelberg.

@@

>>>>

We thank James Davenport (WWU) for releasing his open source spectroscopic
reduction code, \package{PyDIS}, which served as inspiration for the pipeline
used in this work.
This work is based on observations obtained at the MDM Observatory, operated by
Dartmouth College, Columbia University, Ohio State University, Ohio University,
and the University of Michigan.
This work has made use of data from the European Space Agency (ESA)
mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by
the {\it Gaia} Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding
for the DPAC has been provided by national institutions, in particular
the institutions participating in the {\it Gaia} Multilateral Agreement.
This research has made use of the SIMBAD database, operated at CDS, Strasbourg,
France.
This project was developed in part at the 2017 Heidelberg Gaia Sprint, hosted by
the Max-Planck-Institut f\"ur Astronomie, Heidelberg.

@@

\section{Mixture model...}\label{sec:TODO}

% TODO: assumptions:
% - Same velocity dispersion for all ages. Could have an age-dependent
%   dispersion
% -
With the new radial velocity measurements, we construct a statistical model to
determine the probability that a given candidate comoving pair is truly
comoving.
We don't model the positional separation distribution and instead construct a
model purely in velocity-space.
The data are therefore proper motions in the ICRS reference frame, $\bs{\mu}$,
and parallax, $\varpi$, (from the \tgas\ catalog) and barycentric radial
velocity, $v_r$, for each star in a candidate comoving pair; we assume that the
sky position for each star $(\alpha, \delta)$ is known perfectly.
For brevity in the expressions below, we define $\bs{\eta} = (\mu_{\alpha^*},
\mu_\delta, v_r)$, where, e.g., $\bs{\eta}_1$ refers to the proper motion and
radial velocity data for star 1 in the pair.
To simplify some expressions, we additionally ignore any correlations between
the observed parallax and the proper motion components (but keep the covariance
structure between the two proper motion components).

Our model is expressed in terms of the magnitude of the true (unobserved) 3D
velocity difference for each pair, $\psi = \left| \bs{v}_1 - \bs{v}_2 \right|$,
and the (global) fraction of pairs that are comoving, $f$.
For the distribution of true 3D velocity difference magnitudes, $p(\psi \given
f)$, we use a mixture model with two components
\begin{eqnarray}
    p(\psi \given f) &=& f\,p_{\rm co}(\psi) + (1-f)\,p_{\rm field}(\psi)
\end{eqnarray}
where $p_{\rm co}$, the probability distribution function (pdf) of true comoving
pairs, and $p_{\rm field}$, the pdf of randomly paired field stars.\footnote{
This is actually the marginal likelihood under a model in which every candidate
comoving pair has an additional latent parameter: a binary flag, $q_n$, that
specifies which mixture component it belongs too
(\citealt{Hogg:2010,Foreman-Mackey:2014}). We later use this fact to compute the
probability that each pair is truly comoving, or, more precisely, belongs to the
mixture component $p_{\rm co}$).}
The form for both mixture components ($p_{\rm co}$ and $p_{\rm field}$) is taken
to be the probability of the magnitude of the 3D separation of two vectors under
the assumption that the two vectors were drawn independently from an isotropic,
3D Gaussian with dispersion $\sigma$.
That is,
\begin{eqnarray}
    f(\psi \,;\, \sigma) &=& \left[2\sqrt {\pi}\, \sigma^3 \right]^{-1} \,
        \psi^2 \, \ \exp\left[-\frac {\psi^2} {4\, \sigma^2} \right] \\
    \bs{v_1}, \bs{v_2} &\sim& \mathcal{N}(\bs{0}, \sigma^2\,\mat{I}_3) \\
    p_{\rm co}(\psi) &=& f(\psi \,;\, \sigma = 1~\kms)\\
    p_{\rm field}(\psi) &=& f(\psi \,;\, \sigma = \sigma_{\rm field})
\end{eqnarray}
where $\sigma_{\rm field}$ is defined below, $\mathcal{N}(\bs{m},
\mat{\Sigma})$ represents a normal distribution with mean vector $\bs{m}$ and
covariance matrix $\mat{\Sigma}$, and $\mat{I}_3$ is the 3 by 3 identity matrix.

Under this model, the joint probability of the velocity data for a given pair
$\bs{\eta}_1$ and $\bs{\eta}_2$, and the latent parameters $\bs{v}_1, \bs{v}_2,
\psi$, conditional on the unobserved true distances $d_1$ and $d_2$ (we handle
these separately below), is
\begin{equation}
p(\bs{\eta}_1, \bs{\eta}_2, \bs{v}_1, \bs{v}_2, \psi \given d_1, d_2, f) =
    p(\bs{\eta}_1 \given d_1, \bs{v}_1) \,
    p(\bs{\eta}_2, v_{r,2} \given d_2, \bs{v}_2) \,
    p(\bs{v}_1, \bs{v}_2 \given \psi) \, p(\psi \given f)\label{eq:full-ll}
\end{equation}
where
\begin{eqnarray}
    p(\bs{v}_1, \bs{v}_2 \given \psi) &=& \delta(|\bs{v}_1-\bs{v}_2| - \psi) \quad .
\end{eqnarray}

Because we are only interested in inferring the fraction of stars consistent
with being comoving, we marginalize this expression over the true 3D velocities
and velocity difference, obtaining the velocity-marginalized, per-pair
likelihood
\begin{equation}
    p(\bs{\eta}_1, \bs{\eta}_2 \given d_1, d_2, f) =
    \int \dd^3\bs{v}_1 \, \dd^3\bs{v}_2 \, \dd \psi \,
    p(\bs{\eta}_1, \bs{\eta}_2, \bs{v}_1, \bs{v}_2, \psi \given d_1, d_2, f)
    \quad .
\end{equation}
This expression is still conditioned on the true distances, $d_1$ and $d_2$, but
we observe the parallaxes $\varpi_1$ and $\varpi_2$.
To remove the final per-pair latent parameters in this model, we therefore also
have to marginalize the likelihood over these distance parameters to obtain the
velocity-and-distance marginalized likelihood
\begin{equation}
    p(\bs{\eta}_1, \bs{\eta}_2, \varpi_1, \varpi_2 \given f) =
        \int \dd d_1 \, \dd d_2 \,
        p(\varpi_1 \given d_1) \, p(\varpi_2 \given d_2) \,
        p(\bs{\eta}_1, \bs{\eta}_2 \given d_1, d_2, f) \,
        p(d_1) \, p(d_2)
\end{equation}
where we assume a uniform space-density prior for the distances
(\citealt{Bailer-Jones:2015aa}) with a maximum distance of 1~kpc.

The posterior probability over the fraction of true comoving star pairs, $f$, is
then given by
\begin{equation}
    p(f \given \{\bs{\eta}_1, \bs{\eta}_2, \varpi_1, \varpi_2\}_{N}) \propto
        p(f) \, \prod_n^N
        p(\bs{\eta}_{1,n}, \bs{\eta}_{2,n}, \varpi_{1,n}, \varpi_{2,n} \given f)
\end{equation}
where $N$ is the number of candidate comoving star pairs and the index $n$
specifies the data for the $n$th pair.

We approximate all of the marginalization integrals using numerical procedures.
Given distances $d_1$ and $d_2$, we can approximate the velocity marginalization
using Monte Carlo integration:
\begin{equation}
    p(\bs{\eta}_1, \bs{\eta}_2 \given d_1, d_2, f) \approx
    \frac{1}{K} \, \sum^K_k \,
    p(\psi^{(k)} \given f) \quad .
\end{equation}
In this expression, the $k$-th sample $\psi^{(k)}$ is defined as
\begin{eqnarray}
    \psi^{(k)} &=& |\bs{v}_1^{(k)} - \bs{v}_2^{(k)}| \\
    \bs{v}_1^{(k)} &=& \mat{M}(\alpha_1, \delta_1) \cdot
        \left(\begin{array}{c}
            d_1\,\mu_{\alpha^*, 1}^{(k)} \\
            d_1\,\mu_{\delta, 1}^{(k)} \\
            v_{r, 1}^{(k)}
        \end{array}\right)\\
    \left(\begin{array}{c}
        \mu_{\alpha^*, 1}^{(k)} \\
        \mu_{\delta, 1}^{(k)} \\
        v_{r, 1}^{(k)}
    \end{array}\right) &\sim& \mathcal{N}(\bs{\eta}_1, \mat{C}_1)
\end{eqnarray}
where $\mat{C}_1$ is the covariance matrix of the \gaia\ data provided by the
\tgas\ catalog with an additional (uncorrelated) diagonal term added for the
uncertainty in the radial velocity measurement, and the rotation matrix
$\mat{M}(\alpha, \delta)$ transforms from the local, tangent-space coordinates
at $(\alpha, \delta)$ to the ICRS (see, e.g., Section~3 of \citealt{Oh:2017}).
We perform the outer integral over the two distances, $d_1$ and $d_2$, using
Simpson's rule on a 2D grid of distances that, in each dimension, spans the
domain $\left[ (\varpi + 3\sigma_{\varpi})^{-1}, (\varpi -
3\sigma_{\varpi})^{-1} \right]$, where $\sigma_{\varpi}$ is the parallax
uncertainty reported in the \tgas\ catalog for each star.
From convergence tests, we have found that it is sufficient to use 1024 samples
in the Monte Carlo integral and 5 by 5 distance grid points.
